---
title: "Meninge & Parietal Tissue"
subtitle: "Variance"
author: "Kennedi Todd"
date: "09/14/2023"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

# Setup
## Working directory
```{r working_directory}
knitr::opts_knit$set(root.dir = ".")
```

## Libraries
```{r libraries}
library(corrplot)          # corrplot()
library(dplyr)             # %>%
library(edgeR)             # load DGEList object
library(ggcorrplot)        # ggcorrplot()
library(gplots)            # heatmap.2()
library(limma)             # voomWithQualityWeights()
library(mvIC)              # mvForwardStepwise()
library(variancePartition) # fitExtractVarPartModel()
```

## User defined variables
```{r set_variables}
folder <- "both_tissues"
groups <- c("control","meningitis")
group_colors <- c("orange","cornflowerblue")
ann <- "../../refs/chm13v2.0_RefSeq_Liftoff_v5.1.gff3"
```

## Save functions
These functions with help simultaneously save plots as a png, pdf, and tiff 
file.
```{r save_functions}
saveToPDF <- function(...) {
    d = dev.copy(pdf,...)
    dev.off(d)
}
```

## Load data
```{r load_data}
dge.filtered.norm <- readRDS(paste0("../../rObjects/",folder,"_DGEList_filtered_normalized.rds"))
```

# Cell Markers
## Extract metadata
```{r}
# get meta
meta <- as.data.frame(dge.filtered.norm$samples)

# keep variables with all fields known
variables <- c("tissue",
               "group",
               "sex",
               "lib.size",
               "braak",
               "thal",
               "age_at_death",
               "brain_wt",
               "race",
               "RIN",
               "star_unique_mapped")
meta <- meta %>% select(all_of(variables))

# log2 transform lib.size
meta$lib.size <- log2(meta$lib.size)
colnames(meta)[4] <- "log2.lib.size"

# check and change classes
meta$tissue <- factor(meta$tissue)
meta$braak <- factor(meta$braak)
meta$thal <- factor(meta$thal)
meta$race <- factor(meta$race)
meta$RIN <- as.numeric(meta$RIN)
meta$star_unique_mapped <- as.numeric(gsub("%","",meta$star_unique_mapped))
sapply(meta, class)
```

## Markers
- Astrocytes: GFAP, GJA1, CLU, AQP4
- B cells & Plasma Cells: CD19, MS4A1, CD79A, IGHA, SDC1
- Choroid: TTR
- Endothelial: FLT1, PECAM1, VWF \
- Fibroblasts: COL1A1, COL1A2, LUM, DCN \
- Macrophages/Microglia: MRC1, MKI67, C1QB \
- Mast cells: KIT, FCER1A \
- Monocytes/DCs/Neutrophils: ITGAX, CD14, CD209A, S100A9 \
- Neurons: GAD1, GAD2, SNAP25, SYT1, MEG3
- NPCs: SOX2, NES
- NK cells: NKG7
- Oligodendrocytes: MBP
- OPCs: OLIG1, TNR
- Pericytes & SMCs: MYH11, ACTA2, VTN \
- RBCs - HBB, HBA1, HBA2 \
- Schwann cells: MPZ \
- T cells: CD3E, CD3D, CD8A \
```{r}
# choose cell type markers
markers <- c("GFAP", "CD19", "TTR", "VWF", "COL1A2", "MRC1", "KIT", "ITGAX", "SNAP25",
             "NES","NKG7", "MBP", "TNR", "MYH11", "HBB", "MPZ","CD3E")
length(markers)

# check if all markers are in annotation (could have alternative name)
markers[!markers %in% rownames(dge.filtered.norm)]

# get counts
cpm <- round(as.data.frame(cpm(dge.filtered.norm$counts, log = TRUE)), digits = 2)

# filter counts by marker
cpm <- cpm[rownames(cpm) %in% markers,]

# switch cols and rows
cpm <- t(cpm)

# add to meta
all.equal(rownames(meta), rownames(cpm))
meta <- cbind(meta, cpm)
```

# Correlation
## Spearman correlation
- Spearman's correlation can be used to analyzed categorical data, as well as continuous data, because it uses ranks instead of assumptions of normality ().
plotCorrMatrix(cor(info))
MRC1, MKI67, C1QB
```{r CCA_heatmap}
# formula
form <- ~0 + tissue + group + sex + log2.lib.size + braak + thal + 
  age_at_death + brain_wt + race + RIN + star_unique_mapped + NES + MPZ + TNR + 
  KIT + COL1A2 + MRC1 + HBB + CD3E + VWF + MYH11 + CD19 + ITGAX + GFAP + TTR + 
  MBP + NKG7 + SNAP25

# correlation plot
model.matrix(form, data = meta) %>% 
  cor(method = "spearman") %>% 
  ggcorrplot(show.diag = FALSE, type = "lower", lab = TRUE, lab_size = 2)

# save
path <- paste0("../../results/", folder, "/variance/spearman_correlation_heatmap")
saveToPDF(paste0(path, ".pdf"), width = 12, height = 12)
```

## CCA heatmap
```{r}
c <- canCorPairs(form, meta)

corrplot(c,
         type = "lower",
         is.corr = FALSE,
         order = "hclust",
         col = colorRampPalette(c("blue4", "deepskyblue", "white"))(100),
         addCoef.col = TRUE,
         method = "shade",
         tl.col = "black",
         tl.cex = 0.8,
         number.cex = 0.7,
         diag = TRUE)

# save
path <- paste0("../../results/", folder, "/variance/canonical_correlation_analysis_heatmap")
saveToPDF(paste0(path, ".pdf"), width = 12, height = 12)
```

# Variance partition
## Variance model
```{r variance_partition, message=FALSE, eval=FALSE}
# variables to choose from
geneExpr <- as.matrix(dge.filtered.norm$counts)
colnames(meta)

# age is usually a continuous so model it as a fixed effect "age"
# group is categorical, so model them as random effects "(1|group)"
form <- ~ (1|group) + (1|sex) + (1|braak) + (1|thal) + age_at_death + brain_wt + 
  RIN + MPZ + TNR + KIT + COL1A2 + HBB + CD3E + VWF + MYH11 + CD19 + ITGAX + GFAP + TTR + NKG7 + SNAP25

varPart <- fitExtractVarPartModel(geneExpr, form, meta)
vp <- sortCols(varPart)
saveRDS(vp, paste0("../../rObjects/", folder, "_var_par.rds"))
```

## Violins
```{r variance_violins, eval=FALSE}
# plot
plotVarPart(vp)

# save
path <- paste0("../../results/", folder, "/variance/variance_violins")
saveToPDF(paste0(path, ".pdf"), width = 12, height = 6)
```

## Percent bars
```{r variance_percent_bars}
# plot
plotPercentBars(vp[1:10,])

# save
path <- paste0("../../results/", folder, "/variance/variance_gene_percent_bars")
saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
```

## Group var
```{r variance_group}
# sort genes based on variance explained by group
varPart.df <- as.data.frame(vp)
order.varPart.df <- varPart.df[order(varPart.df$group, decreasing = TRUE),]
head(order.varPart.df["group"], 10)
```

## Sex var
```{r variance_group}
# sort genes based on variance explained by group
order.varPart.df <- varPart.df[order(varPart.df$sex, decreasing = TRUE),]
head(order.varPart.df["sex"], 10)
```

## Lib size var
```{r variance_group}
# sort genes based on variance explained by group
order.varPart.df <- varPart.df[order(varPart.df$log2_library_size, decreasing = TRUE),]
head(order.varPart.df["log2_library_size"], 10)
```

## Thal var
```{r variance_group}
# sort genes based on variance explained by group
order.varPart.df <- varPart.df[order(varPart.df$thal, decreasing = TRUE),]
head(order.varPart.df["thal"], 10)
```

## ACTA2 var
```{r variance_group}
# sort genes based on variance explained by group
order.varPart.df <- varPart.df[order(varPart.df$ACTA2, decreasing = TRUE),]
head(order.varPart.df["ACTA2"], 10)
```

## SNAP25 var
```{r variance_group}
# sort genes based on variance explained by group
order.varPart.df <- varPart.df[order(varPart.df$SNAP25, decreasing = TRUE),]
head(order.varPart.df["SNAP25"], 10)
```

# Finding best model
variancePartition::voomWithDreamWeights() is the same as limma::voom(), except that it allows random effects in the formula. limma::voomWithQualityWeights() includes observational and sample specific quality weights whereas voomWithDreamWeights only includes observational quality weights.

## Forward stepwise regression
```{r}
# get matrix of expression data
v <- voomWithDreamWeights(counts = dge.filtered.norm$counts,
                          formula = (~ 0 + group + sex),
                          data = dge.filtered.norm$samples,
                          BPPARAM = BiocParallel::SnowParam(cores),
                          plot = FALSE)
counts <- v$E

# set base formula
base <- ~ (1 | group) + (1 | sex)

# variable we wish to test
variables <- c("log2_library_size",
               "(1|braak)",
               "(1|thal)",
               "age_at_death",
               "brain_wt",
               "RIN",
               "MPZ","TNR","KIT","COL1A2","HBB","CD3E","VWF","MYH11",
               "CD19","ITGAX","GFAP","TTR","NKG7","SNAP25")

# multivariate forward stepwise regression evaluated by multivariate BIC
bestModel <- mvForwardStepwise(exprObj = counts,
                               baseFormula = base,
                               data = meta,
                               variables = variables)
```

# Save
```{r}
all.equal(rownames(meta), rownames(dge.filtered.norm$samples))
dge.filtered.norm$samples <- cbind(dge.filtered.norm$samples, meta)
saveRDS(dge.filtered.norm, paste0("../../rObjects/",folder, "_DGEList_cell_markers.rds"))
```

```{r}
sessionInfo()
```
